A thermodynamic view of dusty protoplanetary disks

### 2017-05-09 linear.tex
home/work sync, typo correction in two-fluid frequency for SI
misc edits to the main text overall

\subsection{Secular gravitational instability}\label{sgi}
We first consider a razor-thin, self-gravitating disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the Dirac delta function and $\Sigma$ is
the total surface density. Now $\epsilon$ and 
$\tepsilon$ refers to the global dust-to-gas ratio and dust 
fraction, respectively. The background disk is
uniform and we neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The linearized dust-gas drag
term is then $-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2
Q$. The thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi = -2\pi G
\Sigma W/\left|k_x\right|$, where $\Sigma$ is the total surface
density.

@@

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks, the distinction between
Eulerian and Lagragian pressure perturbations is unimportant. 
%{\bf but weak growth in this case}
%However
%in that case the gro

In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi \equiv -\sgn(\omega)\arg\left(\Delta P\Delta\rho^*\right)
        = \sgn(\omega)\arg\left(\Delta\rhog^*\Delta\rhod\right),\notag
\end{align*}
%where we have applied the isothermal equation of state for the second equality.
then $\varphi > 0 $ indicates pressure lagging behind density, which
is true for all the cases. In particular, `linB' indicates that growth
rates vanish as $\varphi \to 0$.  
Thus the streaming instability indeed stems  
from the gas pressure evolution lagging behind that of the dust 
density.

@@

\begin{deluxetable*}{llrrrrrr}
  \tablecolumns{8}
  \tablecaption{Selected eigenfrequencies for the streaming
    instability. \label{si_compare}
  }
  \tablehead{
    \colhead{Mode} & 
    \colhead{$\tau_\mathrm{s}\OmK$} &
    \colhead{$\epsilon$} &
     \colhead{$K_{x,z}$} &
      \colhead{$\sigma/\OmK$ (two-fluid)} &
    \colhead{$\sigma/\OmK$ (one-fluid)} &
    \colhead{$\mathcal{W}$ (arbitrary units)} &
      \colhead{$\Delta P$ lag}
  }
\startdata
 linA, \cite{youdin07b} &  $0.1$       & 3.0 & 30    & $-0.3480 +
 0.4190\ii$ & $0.3640 + 0.4249\ii$ & $0.90$ & $30\degr$\\

linB, \cite{youdin07b} & $0.1$        &  0.2 & 6 & $0.4999 +
0.0155\ii$&   $-0.4981 + 0.0054\ii$  & $1.54$ & $1.2\degr$ \\

>>>>

\subsection{Secular gravitational instability}\label{sgi}
We first consider a razor-thin, self-gravitating disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the Dirac delta function and $\Sigma$ is
the total surface density. Now $\epsilon$ and 
$\tepsilon$ refers to the global dust-to-gas ratio and dust 
fraction, respectively. The background disk is
uniform and we neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The linearized dust-gas drag
term is then $-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2
Q$. The thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi(z=0) = -2\pi G
\Sigma W/\left|k_x\right|$.%, where $\Sigma$ is the total surface
%density.

@@

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find in dust-rich disks with 
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. %This
%suggests that for SI in gas-dominated disks, the distinction between
%Eulerian and Lagragian pressure perturbations is unimportant. 
%{\bf but weak growth in this case}
%However
%in that case the gro

In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi \equiv -\sgn(\omega)\arg\left(\Delta P\Delta\rho^*\right)
        = \sgn(\omega)\arg\left(\Delta\rhog^*\Delta\rhod\right),\notag
\end{align*}
%where we have applied the isothermal equation of state for the second equality.
then $\varphi > 0 $ indicates gas density lagging behind dust density,
which is true for all the cases. In particular, `linB' indicates that
growth rates vanish as $\varphi \to 0$.  Thus the streaming
instability is indeed associated with a phase lag between gas and dust
density evolutions.

@@

\begin{deluxetable*}{llrrrrrr}
  \tablecolumns{8}
  \tablecaption{Selected eigenfrequencies for the streaming
    instability. \label{si_compare}
  }
  \tablehead{
    \colhead{Mode} & 
    \colhead{$\tau_\mathrm{s}\OmK$} &
    \colhead{$\epsilon$} &
     \colhead{$K_{x,z}$} &
      \colhead{$\sigma/\OmK$ (two-fluid)} &
    \colhead{$\sigma/\OmK$ (one-fluid)} &
    \colhead{$\mathcal{W}$ (arbitrary units)} &
      \colhead{$\Delta P$ lag}
  }
\startdata
 linA, \cite{youdin07b} &  $0.1$       & 3.0 & 30    & $0.3480 +
 0.4190\ii$ & $0.3640 + 0.4249\ii$ & $0.90$ & $30\degr$\\

linB, \cite{youdin07b} & $0.1$        &  0.2 & 6 & $-0.4999 +
0.0155\ii$&   $-0.4981 + 0.0054\ii$  & $1.54$ & $1.2\degr$ \\

### 2017-05-09 model.tex
misc edits to the main text overall

In the limit of strong dust-gas drag, $\tstop\to0$, the mixture obeys 
the first-order (in $\tstop$) single-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi_\mathrm{tot} - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align}
\citep{laibe14}, where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}. 
\end{align}
For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

For the mixture's temperature evolution, Eq. \ref{tempeq}, $\gamma$ is
the adiabatic index; $\mathcal{H}_\mathrm{drag}$ represents heating
due to dust-gas drag and $\mathcal{H}_\mathrm{eff}$ is an effective
energy source arising from transforming from the two-fluid to
one-fluid description \citep[see][ for details]{laibe14}. We also
include a simple model of radiative cooling,
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star, and 
$G$ is the gravitational constant. The second equality is the  
thin-disk approximation for the stellar potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

@@

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
For later comparison it is convenient to re-write the temperature
evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  + \mathcal{H} 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
where $\mathcal{H}$ collects together dust-drag heating and radiative
cooling.

@@

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature is lowered with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to behave as if it had a temperture
\emph{increasing} away from $z=0$.

\subsection{Effective energy equation}\label{energy_analogy}
Although we have eliminated the need of an evolutionary energy
equation (Eq. \ref{realenergy}) by choosing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective evolutionary energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. We 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Comparing with Eq. \ref{realenergy}, we see that 
Eq. \ref{eff_energy} is the energy equation for an ideal gas of
adiabatic index $\gamma=1$, but now with the imposed temperature
gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ playing the
role of radiative cooling ($\Lambda$ in 
Eq. \ref{realenergy}).

@@

An effective energy equation can also be derived for other 
fixed gas equations of state, such as polytropic gas. We discuss this
in \S\ref{gen_poly}.

@@

%{\bf
\subsection{Physical disk conditions}
%physical disk conditions underwhich the above approx
%(small/well-coupled dust and isothermal gas) is applicable. smaller
%than mm at few tens of au. give dimensional numbers here
%}

The terminal velocity approximation applies to short
stopping times, $\tstop\OmK\ll 1$; while the locally isothermal
approximation applies to short cooling times, $t_\mathrm{cool}\OmK\ll
1$. In typical protoplanetary disk
models such as the Minimum Mass Solar Nebula, $\tstop$ decreases for
smaller particles and/or at smaller radii \citep{youdin11}. However,
$t_\mathrm{cool}$ is generally only small in the outer disk
\citep{lin15,malygin17}. Combining these results, we estimate that our
framework is best suited for mm (or smaller) sized particles at a few
to 10s of AU in protoplanetary disks. However, note that the
isothermal approximation may be relaxed to other fixed equations of
state (see \S\ref{gen_poly}).

@@

In order to have a well-defined stability problem, we formally
take equilibria with $\mathcal{C}\equiv 0$, namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the 
instabilities of interest.

@@

The physical origin for this vertical buoyancy, which exists even in vertically isothermal
disks, is that coupling gas to dust particles increases
the fluid's inertia, but pressure (i.e. restoring) forces are
unchanged. The increased weight of the fluid resists vertical
oscillations and hence there is an associated buoyancy force. However,
if $\tstop\neq0$ so the gas-dust coupling is imperfect, then gas is no
longer `weighed down' by the dust, and can slip past it. Thus finite
drag diminishes the dust-induced buoyancy.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
The condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}.

We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas. {\bf ref for drift} Thus for
instability we need to violate Eq. \ref{dusty_solberg2}. We explore
this below.

\subsubsection{Vertically well-mixed dust}\label{vert_mixed_theory}
In the special case where the dust-to-gas ratio is
uniform in $z$ the LHS of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Thus the disk is \emph{unstable}, because there is no effective vertical 
buoyancy to stabilize vertical motions ($N_z=0$). This situation is 
identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

\subsubsection{Application to protoplanetary disks} 
%In order to apply the stability criteria
%Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} to typical
%protoplanetary disks 
We now consider the specific equilibria described in 
\S\ref{eqm}. We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs. The third term is always destabilizing, but is small in smooth,
thin disks with radial gradients $O(1/r)$,  $h_\mathrm{g}\ll 1$,
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, a result valid for arbitrarily thin dust
layers.

@@

Positive work is done by a fluid parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements), see Fig. \ref{lag_cartoon}
for the case of dusty gas.

@@

A parcel of the isothermal dusty-gas mixture does 
postive work if  
\begin{align*}
-\sgn\left(\omega\right)\imag\left(\Delta\rhog\Delta\rhod^*\right)>0,
\end{align*}
meaning that gas follows dust (Fig. \ref{lag_cartoon}). Instabilities
are thus not possible if the gas does not respond to dust-drag
(i.e. no back-reaction). The pressure-density lag shown in
Fig. \ref{lag_cartoon} is achieved if, after a parcel reaches maximum
total density, its  gas content is still increasing (see $A$ to $B$ in
Fig. \ref{pdv_cartoon}). This in turn implies a sufficiently large
particle flux \emph{out} of the parcel.

>>>>

In the limit of strong dust-gas drag, $\tstop\to0$, the mixture obeys 
the first-order (in $\tstop$) single-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi_\mathrm{tot} - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align}
\citep{laibe14,price15}, where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}. 
\end{align}
For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

For the mixture's temperature evolution, Eq. \ref{tempeq}, $\gamma$ is
the adiabatic index; $\mathcal{H}_\mathrm{drag}$ represents heating
due to dust-gas drag and $\mathcal{H}_\mathrm{eff}$ is an effective
energy source arising from transforming from the two-fluid to
one-fluid description \citep[see][ for details]{laibe14}. We also
include a simple model of radiative cooling,
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{cooling} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star, and 
$G$ is the gravitational constant. The second equality is the  
thin-disk approximation, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

@@

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
For later comparison it is convenient to re-write the temperature
evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  + \mathcal{H}\label{real_pressure}
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
where $\mathcal{H}$ collects together dust-drag heating and radiative
cooling.

@@

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature decreases with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to behave as if it had a temperture
\emph{increasing} away from $z=0$.

\subsection{Effective energy equation}\label{energy_analogy}
Although we have eliminated the true energy
equation (Eq. \ref{tempeq}) by fixing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
Eliminating $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Comparing with Eq. \ref{real_pressure}, we see that 
Eq. \ref{eff_energy} is the energy equation for an ideal gas of
adiabatic index $\gamma=1$, but now with the imposed temperature
gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ playing the
role of radiative cooling ($\Lambda$ in 
Eq. \ref{tempeq}).

@@

%An effective energy equation can also be derived for other 
%fixed gas equations of state, such as polytropic gas. We discuss this
%in \S\ref{gen_poly}.

@@

%{\bf
\subsection{Physical disk conditions}
%physical disk conditions underwhich the above approx
%(small/well-coupled dust and isothermal gas) is applicable. smaller
%than mm at few tens of au. give dimensional numbers here
%}
The above correspondence between isothermal dusty gas and adiabatic
pure gas is derived under the terminal velocity approximation, which applies
to short stopping times, $\tstop\OmK\ll 1$; and the locally isothermal
approximation, which applies to short cooling times, $t_\mathrm{cool}\OmK\ll
1$. In typical protoplanetary disk
models such as the Minimum Mass Solar Nebula, $\tstop$ decreases for
smaller particles and/or at smaller radii \citep{youdin11}. However,
$t_\mathrm{cool}$ is generally only small in the outer disk
\citep{lin15,malygin17}. Combining these results, we estimate the 
correspondence applicable to particles less than mm in size at a few
to 10s of AU in protoplanetary disks. However, note that the
isothermal approximation may be relaxed to other fixed equations of
state (see \S\ref{gen_poly}).

@@

In order to have a well-defined stability problem, we formally
consider equilibria with $\mathcal{C}\equiv 0$. Namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the 
instabilities of interest.

@@

The physical origin for this vertical buoyancy, which exists even in vertically isothermal
disks, is that coupling gas to dust particles increases
the fluid's inertia, but pressure (i.e. restoring) forces are
unchanged. The increased weight of the fluid resists vertical
oscillations and hence there is an associated buoyancy force. However,
if $\tstop\neq0$ so the gas-dust coupling is imperfect, then gas is no
longer `weighed down' by the dust, and can slip past it. Thus finite
drag diminishes the dust-induced buoyancy. This is similar to reducing
gas buoyancy through cooling \citep{lin15}.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
Then the condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}.

We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas. Thus for 
instability we need to violate Eq. \ref{dusty_solberg2}. We explore
this below.

\subsubsection{Vertically well-mixed dust}\label{vert_mixed_theory}
In the special case where the dust-to-gas ratio is
uniform in $z$ the left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Thus the disk is \emph{unstable}, because there is no effective vertical 
buoyancy to stabilize vertical motions ($N_z=0$). This situation is 
identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

\subsubsection{Application to protoplanetary disks} 
%In order to apply the stability criteria
%Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} to typical
%protoplanetary disks 
We now consider the specific equilibria described in 
\S\ref{eqm}. We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, a conclusion valid for arbitrarily thin
dust layers.

%%%%%%%%%%%%%%%%%%

@@

Positive work is done by a fluid parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density, see Fig. \ref{lag_cartoon}
for the case of dusty gas.

@@

A parcel of the isothermal dusty-gas mixture does 
postive work if  
\begin{align*}
-\sgn\left(\omega\right)\imag\left(\Delta\rhog\Delta\rhod^*\right)>0,
\end{align*}
meaning that gas follows dust (Fig. \ref{lag_cartoon}). Instabilities
are thus not possible if the gas does not respond to dust 
(i.e. no back-reaction).% {\bf can we say this means that any effective
%  dust-drag instability require high dust to gas ratios?} 
The pressure-density lag shown in
Fig. \ref{lag_cartoon} is achieved if, just after the total density of a 
parcel maximizes, its gas content
is increasing, which requires a sufficiently large
particle flux \emph{out} of the parcel. See $A$ to $B$ in 
Fig. \ref{pdv_cartoon}.
